Reanalysis Step 7: Missingness Pattern Analysis

### 2025-09-01 reanalysis-7-missingness_patterns.qmd
fix ai idiiocy

```{r identify-groups}
# Get the cohort_id from the first imputed dataset (should be same across all)
included_ids <- imputed_datasets[[1]]$cohort_id

# Identify excluded patients (those in original but not in imputed)
all_ids <- original_data$cohort_id
excluded_ids <- setdiff(all_ids, included_ids)

# Create indicator in original data
original_data$excluded <- original_data$cohort_id %in% excluded_ids

@@

```{r variable-selection}
# Get variables used in outcome and propensity score models
# First, check what columns exist in the processed data
cat("Sample of columns in processed data:\n")
cat(paste(names(original_data)[1:min(30, length(names(original_data)))], collapse = "\n"))
cat("\n...\n")
cat("Total columns:", length(names(original_data)), "\n\n")

@@

**Variables with Substantial Differences (SMD > 0.1):**
```{r}
#| echo: false
if (length(significant_differences) > 0) {
  cat(paste("- ", significant_differences, collapse = "\n"))
} else {
  cat("No variables showed substantial differences (SMD > 0.1)")
}
```

**Lapse Status and Missingness:**
```{r}
#| echo: false
if (!is.na(lapse_or_excluded_vs_included)) {
  cat(sprintf("- Odds ratio of exclusion for lapsed vs not lapsed: %.2f\n", 
              lapse_or_excluded_vs_included))
  if (lapse_or_excluded_vs_included > 1.5) {
    cat("- Lapsed patients were substantially more likely to have missing outcomes\n")
  }
}
```

>>>>

```{r identify-groups}
# Check if cohort_id exists, if not use a unique identifier
id_var <- if("cohort_id" %in% names(original_data)) "cohort_id" else "e_mrn_deidentified"

# Alternative approach: identify excluded based on missing outcome
# Patients were excluded from imputation if they had missing outcomes
if ("outcome_va_vi_binary" %in% names(original_data)) {
  # Use missing outcome as the exclusion criterion
  original_data$excluded <- is.na(original_data$outcome_va_vi_binary)
  log_message("Using missing outcome_va_vi_binary to identify excluded patients")
} else if ("outcome_va_logMAR" %in% names(original_data)) {
  # Use missing outcome VA
  original_data$excluded <- is.na(original_data$outcome_va_logMAR)
  log_message("Using missing outcome_va_logMAR to identify excluded patients")
} else {
  # Fall back to comparing with imputed dataset IDs
  # Get the IDs from the first imputed dataset (should be same across all)
  included_ids <- unique(imputed_datasets[[1]][[id_var]])

  # Get all IDs from original data
  all_ids <- unique(original_data[[id_var]])

  # Identify excluded patients (those in original but not in imputed)
  excluded_ids <- setdiff(all_ids, included_ids)

  # Create indicator in original data (TRUE = excluded, FALSE = included)
  original_data$excluded <- original_data[[id_var]] %in% excluded_ids
  log_message(paste("Using ID comparison to identify excluded patients.",
                   "Original IDs:", length(all_ids),
                   "Included IDs:", length(included_ids)))
}

# Verify the excluded variable has two levels
if (length(unique(original_data$excluded)) < 2) {
  log_message(paste("Warning: All patients are", 
                   ifelse(all(original_data$excluded), "excluded", "included")))
  log_message("Check data processing - there should be both excluded and included patients")
}

@@

```{r variable-selection}
# Get variables used in outcome and propensity score models
# Log what columns exist in the processed data
log_message(paste("Total columns in processed data:", length(names(original_data))))
log_message(paste("Sample columns:", paste(names(original_data)[1:min(10, length(names(original_data)))], collapse = ", ")))

@@

**Variables with Substantial Differences (SMD > 0.1):**
```{r}
#| echo: false
if (length(significant_differences) > 0) {
  for (var in significant_differences) {
    cat("- ", var, "\n")
  }
} else {
  cat("No variables showed substantial differences (SMD > 0.1)\n")
}
```

```{r lapse-missingness-summary}
#| echo: false
# Create summary of lapse status and missingness relationship
if (!is.na(lapse_or_excluded_vs_included)) {
  lapse_summary <- data.frame(
    Measure = c("Odds ratio of exclusion (lapsed vs not lapsed)",
                "Interpretation"),
    Value = c(sprintf("%.2f", lapse_or_excluded_vs_included),
              ifelse(lapse_or_excluded_vs_included > 1.5,
                     "Lapsed patients were substantially more likely to have missing outcomes",
                     "No substantial difference in missingness by lapse status"))
  )
  kable(lapse_summary, 
        caption = "Lapse Status and Missingness Relationship",
        col.names = c("Measure", "Value"))
} else {
  cat("Lapse status and missingness relationship could not be calculated.\n")
}
```
